---
title: "JHopkins Covid Project"
author: "CFT"
date: "2023-03-22"
output: html_document
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning = FALSE, message = FALSE)
```

# Population and Latitude as a model for Covid Cases and Death

Data stored by Johns Hopkins was collected from https://github.com/CSSEGISandData/COVID-19/tree/master/csse_covid_19_data/csse_covid_19_time_series. This data contains the number of covid cases and deaths from January 2020 to March 2023 globally and includes factors Province_State, Country_Region, Lat (lattitude), date, and population. 

This report will analyze the effects of latitude and population on total and rates of covid cases and deaths over time. It is well documented that the further from the equator there is a general trend of greater GDP per capita so the distance from the equator, the absolute value of the latitude, is the true focus. I created a variable called Eq_Dist to represent this. I also created variables deaths per thousand and cases per thousand that reflect the deaths per thousand of a population in a particular country. I collected both global data and US data to model. 

The data is downloaded and cleaned in the sections below.

```{r importcleandata,show_col_types = FALSE}
library(tidyverse)
library(lubridate)
link1 <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"
files <- c("time_series_covid19_confirmed_US.csv",  "time_series_covid19_confirmed_global.csv", "time_series_covid19_deaths_US.csv",  "time_series_covid19_deaths_global.csv")
link1_files <- str_c(link1,files) 
# create data frames of files 
us_covid <- read.csv(link1_files[1],check.names=FALSE)
global_covid <- read.csv(link1_files[2], check.names=FALSE)
us_covidde <- read.csv(link1_files[3],check.names=FALSE)
global_covidde <- read.csv(link1_files[4],check.names=FALSE)
#cleanup of global data
# converting dates from column values to a single row combined with cases or deaths and removing longitude
global_covid <- global_covid %>% pivot_longer(cols = -c(`Province/State`,`Country/Region`,Lat,Long), names_to = "date", values_to = "cases") %>% select(-c(Long))
global_covidde <- global_covidde %>% pivot_longer(cols = -c(`Province/State`,`Country/Region`,Lat,Long), names_to = "date", values_to = "deaths") %>% select(-c(Long))
# joining global covid cases with deaths
global_covid_all <- global_covid %>% full_join(global_covidde) %>% rename(Country_Region = `Country/Region`,Province_State = `Province/State`) %>% mutate(date = mdy(date))
# removing data for times with zero cases and replacing blanks with NA 
global_covid_pos <- global_covid_all %>% filter(cases > 0) %>% mutate_all(na_if,"")
# inputting additional data for population
uid_lookup_url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"
uid <- read_csv(uid_lookup_url) %>% select(-c(Lat,Long_, Combined_Key, code3, iso2, iso3, Admin2)) 
# combining population by Province_State and Country_Region data with global covid data
global_covid_pos <- global_covid_pos %>% left_join(uid, by = c("Province_State","Country_Region")) %>% select(-c(UID,FIPS)) 
# creating Distance from Equator value and removing rows with NAs in Eq_Dist and Population
global_covid_pos <- global_covid_pos %>% mutate(Eq_Dist = round(abs(Lat))) %>% filter(!is.na(Eq_Dist))  %>% filter(!is.na(Population))
#create dataframe of global data grouped by distance from equator
global_by_Eq_Dist <- global_covid_pos %>% group_by (Eq_Dist, date) %>% summarize(cases = sum(cases), deaths = sum(deaths), Population = sum(Population)) %>% mutate(deaths_per_thou = deaths*1000/Population) %>% select(Eq_Dist, date, cases, deaths, deaths_per_thou, Population) %>% ungroup()


#US DATA
# converting columns of dates for covid case data to rows
us_covid <- us_covid %>% pivot_longer(cols = -(UID:Combined_Key), names_to = "date", values_to = "cases") %>% select(Admin2:cases) %>% mutate(date = mdy(date))
# converting columns of dates for covid death data to rows
us_covidde <- us_covidde %>% pivot_longer(cols = -(UID:Population), names_to = "date", values_to = "deaths") %>% select(Admin2:deaths) %>% mutate(date = mdy(date))
# combining US covid case and death data to one data frame
us_all <- us_covid %>% full_join(us_covidde) %>% mutate(Eq_Dist = round(abs(Lat))) %>% filter(!is.na(Eq_Dist))  %>% filter(!is.na(Population))  %>% filter(cases > 0) %>% filter(Population > 0)
# grouping data by distance from equator and adding deaths_per_thou (deaths per thousand) variable
us_by_Eq_Dist <- us_all %>% group_by (Eq_Dist, date) %>% summarize(cases = sum(cases), deaths = sum(deaths), Population = sum(Population)) %>% mutate(deaths_per_thou = deaths*1000/Population) %>% select(Eq_Dist, date, cases, deaths, deaths_per_thou, Population) %>% ungroup()
# adding variables for new cases and new deaths 
us_by_Eq_Dist <- us_by_Eq_Dist %>% mutate(new_cases = cases - lag(cases), new_deaths = deaths - lag(deaths)) %>% mutate(new_cases = coalesce(new_cases,0)) %>% mutate(new_deaths = coalesce(new_deaths,0))



```
Below I will establish several models that attempt to predict deaths from distance to equator using both global data and US data. 


```{r globaldatamodel, echo=FALSE}
#modelss for US Data
gmod1 <- lm(deaths ~ Eq_Dist, global_by_Eq_Dist) 
summary(gmod1)
gmod2 <- lm(deaths_per_thou ~ Eq_Dist, global_by_Eq_Dist) 
summary(gmod2)
gmod3 <- lm(deaths ~ cases + Population + Eq_Dist, global_by_Eq_Dist) 
summary(gmod3)
gmod4 <- lm(deaths_per_thou ~ cases + Population + Eq_Dist, global_by_Eq_Dist)
summary(gmod4)
# adding columns of predicted values based on the models to dataframe of US Covid data grouped by distance from equator 
global_by_Eq_Dist_Pred <- global_by_Eq_Dist %>% mutate(pred1 = predict(gmod1),pred2 = predict(gmod2), pred3 = predict(gmod3),pred4 = predict(gmod4))
```


```{r usdatamodel, echo=FALSE}
#modelss for US Data
mod1 <- lm(deaths ~ Eq_Dist, us_by_Eq_Dist) 
summary(mod1)
mod2 <- lm(deaths_per_thou ~ Eq_Dist, us_by_Eq_Dist) 
summary(mod2)
mod3 <- lm(deaths ~ cases + Population + Eq_Dist, us_by_Eq_Dist) 
summary(mod3)
mod4 <- lm(deaths_per_thou ~ cases + Population + Eq_Dist, us_by_Eq_Dist)
summary(mod4)
# adding columns of predicted values based on the models to dataframe of US Covid data grouped by distance from equator 
us_by_Eq_Dist_Pred <- us_by_Eq_Dist %>% mutate(pred1 = predict(mod1),pred2 = predict(mod2), pred3 = predict(mod3),pred4 = predict(mod4))
```

One can see that distance from the equator (Eq_Dist) alone is not a significant model for deaths in a global setting but does serve as a significant model, according to the estimated probability, for US deaths alone. Distance from the equator alone or with Population and Cases is a significant predictor of deaths per thousand. Below I plotted several of the models for both global and US data. 

## Global Plots 
```{r GlobalModelPlot, echo=FALSE}
#plotting data of deaths by distance from Equator against model of relationship
global_by_Eq_Dist_Pred %>% ggplot() + geom_point(aes(x = Eq_Dist, y = deaths), color = "green", alpha = 0.03) + geom_point(aes(x = Eq_Dist, y = pred1), color = "red")

global_by_Eq_Dist_Pred %>% ggplot() + geom_point(aes(x = Eq_Dist, y = deaths_per_thou), color = "green",alpha = 0.03) + geom_point(aes(x = Eq_Dist, y = pred2), color = "red")
```

## US Plots

```{r USModelPlot, echo=FALSE}
#plotting data of deaths by distance from Equator against model of relationship
us_by_Eq_Dist_Pred %>% ggplot() + geom_point(aes(x = Eq_Dist, y = deaths), color = "blue", alpha = 0.03) + geom_point(aes(x = Eq_Dist, y = pred1), color = "red")

us_by_Eq_Dist_Pred %>% ggplot() + geom_point(aes(x = Eq_Dist, y = deaths_per_thou), color = "blue",alpha = 0.03) + geom_point(aes(x = Eq_Dist, y = pred2), color = "red")
```


## Conclusion and Bias 
There is a statistically significant relationship between distance from Equator and covid deaths per thousand, even with accounting for the other variables cases and population, when using a linear regression model. It was interesting that the relationship between deaths and distance from equator and deaths per thousand and distance from the equator had opposite signs when accounting for population and number of cases.Also there was a negative correlation between deaths per thousand and distance from the equator in the USA but a postivie correlation in the global data. 

There is potential bias in the data for several reasons. First, places with poor health infrastructure, which is typically closer to the equator, may not have been able to accurately collect data on covid related deaths because of lack of testing availability or other access points. Second, many areas could wish to modify their covid death reports for political or economic reasons. Covid restrictions or lack thereof are a potential way to subdue or pacify citizens so governments could want the data to support the policies rather than the other way around.